---
title: "Dry bulk density of surface sediments - DataPrep"
output:
  html_document:
    df_print: paged
---

# Preparations

## Install packages

```{r packages, message=FALSE, warning=FALSE}
rm(list=ls())

library(sf)
library(dplyr)
library(stars)
library(terra)
```


## Define projection and resolution

Projection based on https://projectionwizard.org/ using the AoI.

```{r projection}
crs <- "+proj=laea +lat_0=90 +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs" 
res <- 4000
```


## Define Area of Interest (AoI)

The area of interest is defined by the predicted sediment classes. Rocks and boulders (50) define areas outside the AoI.

```{r aoi}
AoI <- rast("input/GrainSizeReg_folk8_classes_2023-06-28.tif")
AoI[AoI == 11] <- 1
AoI[AoI == 12] <- 1
AoI[AoI == 13] <- 1
AoI[AoI == 20] <- 1
AoI[AoI == 30] <- 1
AoI[AoI == 40] <- 1
AoI[AoI == 50] <- NA
AoI[AoI == 60] <- 1

AoI <- as.polygons(AoI, dissolve = TRUE)
```


# Predictor variables

Potentially relevant predictor variables are loaded and a raster stack is created.

```{r load_predictors}
predictors <- rast("input/predictors_ngb.tif")
grainsize_prob <- extend(rast("input/GrainSizeReg_folk8_probabilities_2023-06-28.tif"), predictors)
predictors <- c(predictors, grainsize_prob)
names(predictors)[38] <- "M"
names(predictors)[39] <- "sM"
names(predictors)[40] <- "mS"
names(predictors)[41] <- "S"
names(predictors)[42] <- "CS"
names(predictors)[43] <- "Mx"
names(predictors)[44] <- "R"
names(predictors)[45] <- "Mos"
```


## Ensure uniform projection

Check if AoI and predictors have the defined projection. Re-project if this is not the case.

```{r uniform_proj}
if (st_crs(AoI)$proj4string != crs) {
  AoI <- st_transform(AoI, crs)
}

if (crs(predictors) != crs) {
  predictors <- project(predictors, crs, res = res)
}
```


## Crop predictors to AoI

```{r crop_predictors}
predictors <- crop(mask(predictors, AoI, touches = FALSE), AoI)
plot(predictors)
```


## Create a fishnet based on predictor raster

```{r fishnet}
min_ext <- sum(predictors)
min_ext[min_ext == 0] <- NA
fishnet <- st_as_sf(stars::st_as_stars(min_ext), as_points = FALSE, merge = FALSE)
fishnet$ID <- 1:nrow(fishnet)

#write_sf(fishnet, dsn = "input", layer = "fishnet", driver = "ESRI Shapefile")
```


# Response variable

## Load response

```{r load_response}
DBD_data <- read.csv("input/DBD_2023-07-21.csv", header = TRUE, sep = ",")
summary(DBD_data)

names(DBD_data)[5] <- "DBD"

DBD_data <- subset(DBD_data, Event.Label != "R1689") # Removal of a suspicious station

# Convert to sf
DBD <- st_as_sf(DBD_data, coords = c("Longitude", "Latitude"), crs = "+proj=longlat +datum=WGS84 +no_defs")

summary(DBD)
```


## Ensure uniform projection

Check if the response variable has the defined projection. Re-project if this is not the case.

```{r uniform_proj_oc}
if (st_crs(DBD)$proj4string != crs) {
  DBD <- st_transform(DBD, crs)
}
```


## Sample depth intervals

Define breaks in sample depth to create depth intervals. Note that depth is given in cm and as a positive number here.

It has become an unofficial standard to use the upper 10 cm of the sediment column.

```{r depth_intervals}
brks <- c(0, 10)
DBD0 <- subset(DBD, Obs_depth_cm >= brks[1] & Obs_depth_cm < brks[2])
```


Alternatively, standard depth intervals as in terrestrial soil mapping could be used. These are 0 - 5 cm, 5 - 15 cm, 15 - 30 cm, 30 - 60 cm, 60 - 100 cm, and 100 - 200 cm.

```{r depth_intervals_standard}
#brks <- c(0, 5, 15, 30, 60, 100)
#DBD0 <- subset(DBD, Obs_depth_cm >= brks[1] & Obs_depth_cm < brks[2])
#DBD5 <- subset(DBD, Obs_depth_cm >= brks[2] & Obs_depth_cm < brks[3])
#DBD15 <- subset(DBD, Obs_depth_cm >= brks[3] & Obs_depth_cm < brks[4])
#DBD30 <- subset(DBD, Obs_depth_cm >= brks[4] & Obs_depth_cm < brks[5])
#DBD60 <- subset(DBD, Obs_depth_cm >= brks[5] & Obs_depth_cm < brks[6])
```


## Centroids

Centroids are calculated to average all measurements of DBD within one grid cell and the relevant depth interval.

```{r centroids_surface}
DBD_centroids <- st_join(fishnet, DBD0)
DBD_centroids <- na.omit(DBD_centroids)
DBD_centroids <- DBD_centroids %>% group_by(ID) %>% summarize(DBD = mean(DBD, na.rm = TRUE))
DBD0 <- st_centroid(DBD_centroids)
```

... or alternatively for standard depth intervals:

```{r centroids_standard_intervals}
#DBD_centroids <- st_join(fishnet, DBD0)
#DBD_centroids <- na.omit(DBD_centroids)
#DBD_centroids <- DBD_centroids %>% group_by(ID) %>% summarize(DBD = mean(DBD, na.rm = #TRUE))
#DBD0 <- st_centroid(DBD_centroids)

#DBD_centroids <- st_join(fishnet, DBD5)
#DBD_centroids <- na.omit(DBD_centroids)
#DBD_centroids <- DBD_centroids %>% group_by(ID) %>% summarize(DBD = mean(DBD, na.rm = #TRUE))
#DBD5 <- st_centroid(DBD_centroids)

#DBD_centroids <- st_join(fishnet, DBD15)
#DBD_centroids <- na.omit(DBD_centroids)
#DBD_centroids <- DBD_centroids %>% group_by(ID) %>% summarize(DBD = mean(DBD, na.rm = #TRUE))
#DBD15 <- st_centroid(DBD_centroids)

#DBD_centroids <- st_join(fishnet, DBD30)
#DBD_centroids <- na.omit(DBD_centroids)
#DBD_centroids <- DBD_centroids %>% group_by(ID) %>% summarize(DBD = mean(DBD, na.rm = #TRUE))
#DBD30 <- st_centroid(DBD_centroids)

#DBD_centroids <- st_join(fishnet, DBD60)
#DBD_centroids <- na.omit(DBD_centroids)
#DBD_centroids <- DBD_centroids %>% group_by(ID) %>% summarize(DBD = mean(DBD, na.rm = #TRUE))
#DBD60 <- st_centroid(DBD_centroids)
```


# Save files for use in main process

```{r save_predictors}
writeRaster(predictors, "input/predictors.tif", overwrite = TRUE)
```


```{r save_0-10cm}
write_sf(DBD0, dsn = "input", layer = paste0("DBD0_", Sys.Date()), driver = "ESRI Shapefile")
```


```{r save_depth_intervals}
#write_sf(DBD0, dsn = "input", layer = paste0("DBD0_", Sys.Date()), driver = "ESRI Shapefile")
#write_sf(DBD5, dsn = "input", layer = paste0("DBD5_", Sys.Date()), driver = "ESRI Shapefile")
#write_sf(DBD15, dsn = "input", layer = paste0("DBD15_", Sys.Date()), driver = "ESRI Shapefile")
#write_sf(DBD30, dsn = "input", layer = paste0("DBD30_", Sys.Date()), driver = "ESRI Shapefile")
#write_sf(DBD60, dsn = "input", layer = paste0("DBD60_", Sys.Date()), driver = "ESRI Shapefile")
```


## Save session info

```{r save_session_info}
sessionInfo <- sessionInfo()
save(sessionInfo, file = "sessionInfo_DataPrep.Rdata")
rm("sessionInfo")
```


## Save global environment

```{r save_global_env}
save.image(file = "globEnv_DataPrep.RData")
```

